---
title: "Replication Code"
format: html
editor: visual
---

### Intro

All quoted portions in this document are from the following paper: <https://onlinelibrary.wiley.com/doi/10.1002/ajpa.24665>

This paper includes two distinct analyses, one which includes humans and one which does not. The processes are identical between these groups, so I have elected to only do the non-human analysis for this project in order to make differences easier to see.

### Citation

Lewton, Kristi L. “An Integrative Approach to Examining the Influences of Size, Phylogeny, and Locomotion on Os Coxae Shape Variation in Primates.” *American Journal of Biological Anthropology*, vol. 183, no. 3, Mar. 2024, p. e24665. *DOI.org (Crossref)*, https://doi.org/10.1002/ajpa.24665.

### Setup

```{r, include = FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/isabe/OneDrive/Documents/lewton-replication") 
```

```{r}
#| warning: FALSE 
#| outcome: FALSE  

#Packages  
library(geomorph) 
library(geiger) 
library(tidyverse)  
library(ggplot2) 
library(curl) 
library(psych) 
library(Morpho)
library(colorspace)
```

```{r}
#File Creation  
#NHP Landmarks 
landmarks.raw <- read.csv("NHP landmarks.csv", header = TRUE, row.names = 1, stringsAsFactors = FALSE)
View(landmarks.raw)  

#NHP Labels 
labels <- read.csv("NHP labels.csv", header=TRUE, row.names=1, stringsAsFactors = FALSE) 
labels <- as.matrix(labels)  

#Wireframe Map  
#tells which landmarks to draw lines/calc distances between 
wireframe <- as.matrix(read.csv("wireframe.csv", header=FALSE)) 

#Species Names by specimen 
species <-read.csv("NHP specimen taxon locomotion list.csv", header=TRUE, row.names = 1) 
View(species)
```

```{r} 
# Data Cleaning  
#1. Reflect all left-side data so all specimens are right-side  
#Requires labels and landmarks.raw files to be in the same order and same length    

for(i in 1:nrow(labels)) {   
  if(labels[i,8] == "L") { #checks if L/R for each     
    landmarks.raw[i, ] <- (-1)*(landmarks.raw[i, ]) #multiplies all coordinates by -1 if it is L   
  } 
}  

#2. Create coordinates for each spec, each landmark is a row as [X, Y, Z] 
landmarks <- arrayspecs(landmarks.raw, 20, 3)  

#3. Put all the data into one list 

d <- list("landmarks" = landmarks, "labels" = labels, "wireframe" = wireframe) 

#Datasets: d, landmarks, landmarks.raw, labels, wireframe
```

### General Procrustes Analysis

```{r}
gpa <- gpagen(d$landmarks, max.iter=NULL)
summary(gpa)
#20 fixed landmarks
#0 semilandmarks (sliders)
#3-dimensional landmarks
#5 GPA iterations to converge

#Notes
names(gpa)
#$coords refers to the coordinates of each of 10(?) landmarks for each species
#$Csize holds the centroid size calculated for each species based on cumulative distances between each landmark
```

```{r}
#1. Compute shape means by species
coords <- two.d.array(gpa$coords)
species <- species$taxon

shape.means <- rowsum(coords, species)/as.vector(table(species))


#2. Compute centroid size by species
cent.means <-rowsum(gpa$Csize, species)/as.vector(table(species))

#3. 
means <-read.delim("NHP taxon locomotion list means.csv", sep=",", header=TRUE)
row.names(means)<-means$taxon
```

### Procrustes ANOVA - sex and species

"To examine the potential effects of sex differences on os coxae shape, a Procrustes analysis of variance was performed on the Procrustes coordinates...with sex and species as factors."

"The non-phylogenetic models were tested using the procD.lm function in the “geomorph” R package."

```{r}
#Creating a data frame to put into geomorph  

sex_vector <- labels$sex  
gdf <- geomorph.data.frame(shape  = coords,   sex    = sex_vector,    species = species)

m1 <- procD.lm(shape ~ as.factor(sex) * as.factor(species),
                data   = gdf,   
                iter   = 999,
                seed = NULL,
                RRPP = TRUE)
                   
summary(m1)
```

"This analysis demonstrated differences in Procrustes coordinates related to species (F = 48.45, df = 59, p = 0.001), sex (F = 12.97, df = 2, p = 0.001), and the interaction of species and sex (F = 1.63, df = 72, p = 0.001)."

My analysis found different F statistics, but the same pattern of significance regarding sex, species, and their interaction. This might be partially a result of the difference in degrees of freedom for the species term, which was 59 in the original report and 58 in mine. The source of this difference is likely related to the following warning I received while running the model: "Because variables in the linear model are redundant, the linear model design has been truncated (via QR decomposition)." I also notice that the td object has 59 tips and 58 internal nodes, so perhaps my dfs are pulling from that instead?

[sex] F = 6.41, df = 2, p = 0.001

[species] F = 7.81, df = 58, p = 0.001

[sex:species] F = 1.39, df = 71, p = 0.001

I also ran this same analysis with 10,000 iterations, since the number was not specified in the publication, and this is more standard for publishing, compared with the 1000 minimum I used initially.

```{r}
#| warning: FALSE

m1.10k <- procD.lm(shape ~ as.factor(sex) * as.factor(species),
                data   = gdf,   
                iter   = 9999,
                seed = NULL,
                RRPP = TRUE)
                   
summary(m1.10k)
```

The only meaningful difference when more iterations were performed was that the significance of the interaction term was slightly higher (p = 0.0106).

Because these terms were significant, the following analyses were performed on combined-sex and separated sex samples.

###Phylogeny, Locomotion & Body Size

"To examine the effects of locomotion, body size, and phylogeny on os coxae shape... species means were calculated and phylogenetic analysis of covariance (ANCOVA) models were analyzed... Two sets of models were constructed: one that did not incorporate phylogeny and one that did.

The following 2 chunks organize the existing data so that further analysis can be done on the species level, rather than the individual specimen level.

```{r}
#1. Make list of locomotion by species
spec.means <- read.delim("NHP taxon locomotion list means.csv", sep = ",", header = TRUE)
row.names(spec.means) <- spec.means$taxon #set row names to the taxon variable

#2. Make matrixes that can go into the tree
loco <- as.matrix(spec.means$loco)
row.names(loco)<-spec.means$taxon #same row values as above
View(loco)

#3. Call phylogeny
#need to use {geiger}
tree <- read.tree(file = "Springer 2012 tree.txt")
plot(tree, cex = 0.5) #just a check, not pretty! 
#length(tree$tip.label) #tree is class phylo, 372 represented species
```

```{r}
#using geiger package to pair up the tree and the data, "prunes" branches that aren't in the set

td <- treedata(tree, shape.means, sort = TRUE, warning = TRUE)
shape.data <- td$data

td <- treedata(tree, cent.means, sort = TRUE, warning = TRUE)
size.data <- td$data

td <- treedata(tree, loco, sort = TRUE, warning = TRUE)
loco.data <- td$data

#the above prunes a bunch of tips to create three versions of the tree based on what is or is not available from our data sets (shape, size, and locomotion
plot(td$phy, cex = 0.5, color = td$data[,1])
```

### PGLS - Phylogenetic Models

"Concordance between the tree topology and data observations was ensured using the treedata function in the “geiger” R package (Pennell et al., 2014), and distance-based phylogenetic least squares analysis was conducted using the procD.pgls function in the “geomorph” R package (Collyer & Adams, 2018, 2021)."

```{r}
#Is shape meaningfully predicted by body size + locomotion? 

#1. Build geomorph dataframe
gdf2 <- geomorph.data.frame(shape  = shape.data,  size    = size.data,  loco = loco.data, phy = td$phy)

#hist(log(size.data))
#size is more normal when logged- I also ran it without logging and the significance values were the same FYI 

m2 <- procD.pgls(
  shape ~log(size) + loco,
  phy    = td$phy,
  data   = gdf2,
  iter   = 999,
)

#1000 permutations, using GLS method
summary(m2)

#          Df      SS      MS     Rsq      F       Z Pr(>F)  
#log(size)  1  0.3962 0.39624 0.03465 1.9889  1.7167  0.037 *
#loco       6  0.8791 0.14651 0.07687 0.7354 -1.1524  0.879  
#Residuals 51 10.1604 0.19922 0.88848                        
#Total     58 11.4357   
```

"Neither size nor locomotion has a significant effect on non-human primate os coxae shape while accounting for phylogeny (Table 4)."

I found significant effect from log(size) but not from locomotion in this model. This may have to do with how size was defined and manipulated in earlier data cleaning, or because I used log transformation.

### Phylogenetic Signal

"Degree of phylogenetic signal (i.e., the tendency for related taxa to exhibit similar phenotypic trait values, Blomberg et al., 2003) in os coxae shape was assessed using the multivariate K statistic (Kmult, Adams, 2014, implemented in the “geomorph” R package)."

A K of 0 would indicate no influence of phylogenetic relatedness to the outcome, a K of >1 would mean any variance observed in the outcome is likely to be a result of relatedness.

```{r}
signal.shape <- physignal(shape.data, td$phy, iter = 999)
summary(signal.shape)
#K = 0.2943, p = 0.001

signal.size <- physignal(size.data, td$phy, iter = 999)
summary(signal.size)
#K = 1.2334, p = 0.001
```

"There is phylogenetic signal in non-human primate os coxae shape (Kmult = 0.29, p = 0.001) and size (K = 1.23, p = 0.001)."

Phylogenetic signal analysis found that size was most likely a factor of relatedness, but shape was not meaningfully attached to phylogenetic relationships, suggesting there are other factors affecting os coxa shape.


### Non-Phylogenetic Model

```{r}
gdf3 <- geomorph.data.frame(shape  = shape.means,  size    = cent.means,  loco = loco)

m3 <- procD.lm(
  shape ~ log(size) + loco,
  data   = gdf3,
  iter   = 999,
  effect.type = "F"
)
summary(m3)
#log size and loco are both significant
#log(size): F = 39.7677, p = 0.001
#loco: F = 5.6358, p = 0.001
```

"The linear model that does not include phylogeny demonstrates that both size and locomotion have significant effects on nonhuman primate os coxae shape (size R2 = 0.32, F = 39.8, p = 0.001, locomotion R2 = 0.27, F = 5.6, p = 0.001, Table 3)."

In this instance, I was able to accurately replicate the data reported in the paper, with the same R-squared, F, and p values appearing in my model summary. This summary indicates significance of both size and locomotion, which means post-hoc pairwise testing is required for locomotion groups.

### Post-Hoc Testing

```{r}
m3pw <- pairwise(
  fit         = m3,
  fit.null = NULL,
  groups      = gdf3$loco,
  covariate = NULL
)
summ <- summary(m3pw, test.type = "dist", confidence = 0.95)
summ$pairwise.tables$P
```

"Pairwise comparisons of locomotor groups demonstrate that slow climbers are significantly different from all other locomotor groups; suspensory taxa differ from AQ, AQ/TQ, and TQ; and AQ/leap differ from AQ and VCL (Table 7)."

My pairwise test demonstrates that there are statistically significant differences between os coxa shape between numerous pairs of locomotions. When blended categories, species that use AQ and another strategy, are excluded, there are significant differences between all pairs! This is not consistent with what is reported in the paper, though some insights (slow climbers being different than all others) are retained. I am not sure where this distinction comes from, especially because the model itself appeared the same.

### PCA

```{r}
#Using full dataset from above containing wireframe, labels, and landmarks
#d <- list("landmarks" = landmarks, "labels" = labels, "wireframe" = wireframe)

#taxon and locomotion
taxon <- as.factor(d$labels[,11])
locom <- as.factor(d$labels[, 10])

pca.gdf <- geomorph.data.frame(gpa, taxon = taxon, loco = locom)
new.coords <- coords.subset(A = pca.gdf$coords, group = taxon)

spmean <- lapply(new.coords, mshape)
spmean <- simplify2array(spmean, higher=TRUE)

PCA.phylo <- gm.prcomp(spmean, phy=td$phy)
summary(PCA.phylo)

plot(PCA.phylo)
```

"The first principal component accounts for 42% of the variation, while the second principal component accounts for 13% of the variation (Figure 2). Centroid size is strongly correlated with and explains 62% of the variation in principal component 1 (PC1; r = −0.79, p < 0.0001)."

In my analysis, PC1 explains 42% of the variation in landmark distance (our proxy of shape) in this analysis, and PC2 accounts for 14.3% of the variation. Together, these two plotted together will explain 56.3% of variation. I imagine that centroid size is correlated in this case as well, but attempts to run a correlation test were not successful, I think because of problems with my object titles and organization that would take too long to troubleshoot. All in all, my PCA looks very similar to Lewton's.

### Visualization

"Differences in os coxae shape among taxa were explored using a phylomorphospace approach implemented in the “geomorph” R package (Adams et al., 2021; Baken et al., 2021)."

```{r}
data<-read.delim("NHP taxon locomotion list means.csv",sep=",",header=TRUE, row.names=1)
row.names(data)
gp<-factor(data$loco)

q8<-qualitative_hcl(8, palette= "Dark 3")
palette(q8)
```


```{r}
plot(PCA.phylo, flip=1,axis1=1, axis2=2, pch=21, col="black", bg=gp, cex=1.5, bty="L", phylo = TRUE, phylo.par = list(tip.labels=TRUE, node.labels=FALSE, anc.states=FALSE, edge.color="grey", node.cex=0, tip.txt.cex=0.5, tip.txt.adj=c(-0.15,0.2)))

#tried making the wireframe shapes but couldn't figure out what to do after this? oh well! 
ref<-mshape(gpa$coords)
#note: we want to flip axis one so that Fig 3 and Fig 2 have the same layout
PC1max <- plotRefToTarget(ref,PCA.phylo$shapes$shapes.comp1$min, method="points", links=wireframe)
PC1min <- plotRefToTarget(ref,PCA.phylo$shapes$shapes.comp1$max, method="points", links=wireframe)
PC2min <- plotRefToTarget(ref,PCA.phylo$shapes$shapes.comp2$min, method="points", links=wireframe)
PC2max <- plotRefToTarget(ref,PCA.phylo$shapes$shapes.comp2$max, method="points", links=wireframe)

```
